---
params:
    data_file: "~/Downloads/CTG_PBRANT/PBRANT_ctg_viability_data.Rds"
title: "`r basename(dirname(params$data_file))`"
author: "Andrew Boghossian"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, include = F, warning = F, error = F, message = F,
                      fig.width = 12)
library(plyr)
library(tidyverse)
library(ggthemes)
library(magrittr)
library(future.apply)
library(psych)
theme_set(theme_bw(base_size = 12))
suppressMessages(source("src/helperFunctions.R"))
suppressMessages(source("src/make_dose_curves.R", chdir = T))

project <- word(basename(dirname(params$data_file)), 2, -1, sep = fixed("_"))
```

```{r load data}
normalized_data <- read_rds(params$data_file)
if (all(is.na(normalized_data$plate_map_name)) &
    "Source Barcode" %in% colnames(normalized_data)) {
  normalized_data %<>% dplyr::mutate(plate_map_name = `Source Barcode`)
}
```

# Introduction

This is a QC of the `r project` CTG plates.

# Luminescence and viability distributions {.tabset .tabset-pills}

## Luminescence distributions

```{r luminescence, include = T, fig.height = 10}
normalized_data %>% 
  ggplot(aes(x = lum, color = pert_type)) + 
  geom_density() +
  geom_histogram(aes( y = ..density.., fill = pert_type), alpha = .3, position = 'identity') + 
  scale_x_log10() +
  labs(x = "Luminescence", y = "Density", color = "Type") +
  ggtitle('Distribution of luminescence, separated by treatment type') +
  scale_color_tableau(guide = "none") + scale_fill_tableau(name = "pert_type") +
  facet_grid(plate_map_name + replicate ~ ccle_name)
```

## Viability distributions (poscon normalized fold-change)

```{r, include = T, fig.height = 10}
normalized_data %>% 
  ggplot(aes(x = viability, color = pert_type)) + 
  geom_density() +
  geom_histogram(aes( y = ..density.., fill = pert_type), alpha = .3, position = 'identity') + 
  scale_x_log10() +
  labs(x = "Viability (poscon normalized)", y = "Density", color = "Type") +
  scale_color_tableau(guide = "none") + scale_fill_tableau(name = "pert_type") +
  ggtitle('Distribution of viability, separated by treatment type') +
  facet_grid(plate_map_name + replicate ~ ccle_name)
```

## Viability distributions (fold-change)

```{r, include = T, fig.height = 10}
normalized_data %>% 
  ggplot(aes(x = 2^log_fc, color = pert_type)) + 
  geom_density() +
  geom_histogram(aes( y = ..density.., fill = pert_type), alpha = .3, position = 'identity') + 
  scale_x_log10() +
  labs(x = "Viability", y = "Density", color = "Type") +
  scale_color_tableau(guide = "none") + scale_fill_tableau(name = "pert_type") +
  ggtitle('Distribution of viability, separated by treatment type') +
  facet_grid(plate_map_name + replicate ~ ccle_name)
```

# SSMDs

```{r calc ssmd}
ssmds <- normalized_data %>%
  dplyr::group_by(arp_barcode, ccle_name, plate_map_name, replicate) %>%
  dplyr::summarise(ssmd = calculate_ssmd_robust(log_fc[pert_type == 'poscon'],
                                                log_fc[pert_type == 'negcon'])) %>%
  dplyr::ungroup()
```

```{r ssmd plot, include = T}
ssmds %>%
  DT::datatable()

ssmds %>%
  ggplot(aes(x = ssmd, fill = paste(word(ccle_name, 1, sep = fixed("_")), plate_map_name, replicate))) + 
  geom_histogram() + 
  labs(x = "SSMD", y = "Count", fill = "Cell Line Replicate") +
  scale_y_continuous(limits = c(0, NA)) +
  geom_vline(aes(xintercept = -2), linetype = "dashed")
```

# Replicate correlations {.tabset .tabset-pills}

## Viability (poscon normalized)
```{r, include = T, fig.height=8}
normalized_data %>%
  dplyr::filter(pert_type != "treatment" | !is.na(dose)) %>%
  tidyr::pivot_wider(id_cols = c("ccle_name", "pert_type", "dose", "broad_id"),
                     names_from = c("plate_map_name", "replicate"),
                     values_from = "viability",
                     values_fn = median) %>%
  dplyr::select(-ccle_name, -dose, -pert_type, -broad_id) %>%
  psych::pairs.panels(ellipses = F, hist.col = tableau_color_pal()(1))
  
```

## Viability (log fold-change)
```{r, include = T, fig.height=8}
normalized_data %>%
  dplyr::filter(!is.na(dose)) %>%
  tidyr::pivot_wider(id_cols = c("ccle_name", "pert_type", "dose", "broad_id"),
                     names_from = c("plate_map_name", "replicate"),
                     values_from = "log_fc",
                     values_fn = median) %>%
  dplyr::select(-ccle_name, -dose, -pert_type, -broad_id) %>%
  psych::pairs.panels(ellipses = F, hist.col = tableau_color_pal()(1))
  
```

# Does viability decrease with dose?

## Spearman correlation

We can use spearman correlation to measure the monotonicity of dose vs viability. 

Negative correlations indicate dose-response behavior.

```{r correlations, include = TRUE}
spearman_correlations <- normalized_data %>%
  dplyr::group_by(arp_barcode, broad_id, ccle_name, plate_map_name, replicate) %>%
  dplyr::filter(length(unique(dose)) > 3) %>%
  dplyr::summarise(spearman = cor(viability, dose, method = 'spearman', use = 'p')) %>%
  dplyr::ungroup()

ggplot(spearman_correlations, aes(y = spearman, x = replicate)) + 
  geom_boxplot(aes(fill = replicate)) + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme(axis.text.x = element_text(angle = 45)) + 
  scale_fill_tableau(guide = "none") +
  facet_grid(plate_map_name ~ ccle_name, scales = 'free') +
  labs(x = "Replicate", y = "Spearman Correlation")
```

## Treatment dose trends

**Note:** these plots are limited to the range $[-5, 150]$ and exclude some outlying points (particularly negative values) on worse performing plates.

```{r, include = T, fig.height = 160}
normalized_data %>%
  dplyr::filter(pert_type == "treatment", !is.na(dose)) %>%
  ggplot(aes(x = dose, y = viability, color = arp_barcode)) +
  geom_line() + geom_point() +
  scale_x_log10() +
  facet_wrap(broad_id ~ ccle_name, scales = "free_x", ncol = 6) +
  labs(color = "Plate") +
  ylim(c(-5, 150))
```
