---
params:
    data_file: "~/Desktop/MTS014_CTG/MTS014_ctg_viability_data.Rds"
title: "`r basename(dirname(params$data_file))`"
author: "Andrew Boghossian"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, include = F, warning = F, error = F, message = F,
                      fig.width = 10)
library(plyr)
library(tidyverse)
library(ggthemes)
library(magrittr)
library(future.apply)
library(psych)
theme_set(theme_bw(base_size = 12))
suppressMessages(source("src/helperFunctions.R"))
suppressMessages(source("src/make_dose_curves.R", chdir = T))

project <- word(basename(dirname(params$data_file)), 1, -2, sep = fixed("_"))
```

```{r load data}
normalized_data <- read_rds(params$data_file)
if (all(is.na(normalized_data$plate_map_name)) &
    "Source Barcode" %in% colnames(normalized_data)) {
  normalized_data %<>% dplyr::mutate(plate_map_name = `Source Barcode`)
}
```

# Introduction

This is a QC of the `r project` CTG plates.

# Luminescence and viability distributions

## Luminescence distributions

```{r luminescence, include = T, fig.width = 10, fig.height = 10}
normalized_data %>% 
  ggplot(aes(x = lum, color = pert_type)) + 
  geom_density() +
  geom_histogram(aes( y = ..density.., fill = pert_type), alpha = .3, position = 'identity') + 
  scale_x_log10() +
  labs(x = "Luminescence", y = "Density", color = "Type") +
  ggtitle('Distribution of luminescence, separated by treatment type') +
  scale_color_tableau(guide = "none") + scale_fill_tableau(name = "pert_type") +
  facet_grid(plate_map_name + replicate ~ ccle_name)
```

## Control luminescence distributions

```{r, include=T, fig.width=10, fig.height=10}
normalized_data %>%
  dplyr::filter(pert_type != "treatment") %>%
  ggplot(aes(x = lum)) + 
  geom_histogram(aes( y = ..density..), alpha = .5, position = 'identity') + 
  labs(x = "Luminescence", y = "Density", color = "Type") +
  ggtitle('Distribution of luminescence, separated by treatment type') +
  facet_grid(plate_map_name + replicate ~ ccle_name + pert_type, scales = "free")
```

## Viability distributions

```{r viability, include = T, fig.width = 10, fig.height = 10}
normalized_data %>% 
  ggplot(aes(x = viability, color = pert_type)) + 
  geom_density() +
  geom_histogram(aes( y = ..density.., fill = pert_type), alpha = .3, position = 'identity') + 
  scale_x_log10() +
  labs(x = "Viability", y = "Density", color = "Type") +
  scale_color_tableau(guide = "none") + scale_fill_tableau(name = "pert_type") +
  ggtitle('Distribution of viability, separated by treatment type') +
  facet_grid(plate_map_name + replicate ~ ccle_name)
```

# SSMDs

```{r calc ssmd}
ssmds <- normalized_data %>%
  dplyr::group_by(arp_barcode, ccle_name, plate_map_name, replicate) %>%
  dplyr::summarise(ssmd = calculate_ssmd_robust(log_fc[pert_type == 'poscon'],
                                                log_fc[pert_type == 'negcon'])) %>%
  dplyr::ungroup()
```

```{r ssmd plot, include = T}
ssmds %>%
  DT::datatable()

ssmds %>%
  ggplot(aes(x = ssmd, fill = paste(word(ccle_name, 1, sep = fixed("_")), plate_map_name, replicate))) + 
  geom_histogram() + 
  labs(x = "SSMD", y = "Count", fill = "Cell Line Replicate") +
  scale_fill_tableau(palette = "Tableau 20") +
  scale_y_continuous(limits = c(0, NA)) +
  geom_vline(aes(xintercept = -2), linetype = "dashed")
```

# Replicate correlations
```{r, include = T, fig.height=8}
normalized_data %>%
  dplyr::filter(pert_type != "treatment" | !is.na(dose)) %>%
  tidyr::pivot_wider(id_cols = c("ccle_name", "pert_type", "dose", "broad_id"),
                     names_from = c("plate_map_name", "replicate"),
                     values_from = "viability",
                     values_fn = median) %>%
  dplyr::select(-ccle_name, -dose, -pert_type, -broad_id) %>%
  psych::pairs.panels(ellipses = F, hist.col = tableau_color_pal()(1))
  
```

# Does viability decrease with dose?

## Spearman correlation

We can use spearman correlation to measure the monotonicity of dose vs viability. 

Negative correlations indicate dose-response behavior.

```{r correlations, include = TRUE}
spearman_correlations <- normalized_data %>%
  dplyr::group_by(arp_barcode, broad_id, ccle_name, plate_map_name, replicate) %>%
  dplyr::filter(length(unique(dose)) > 3) %>%
  dplyr::summarise(spearman = cor(viability, dose, method = 'spearman', use = 'p')) %>%
  dplyr::ungroup()

ggplot(spearman_correlations, aes(y = spearman, x = replicate)) + 
  geom_boxplot(aes(fill = replicate)) + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme(axis.text.x = element_text(angle = 45)) + 
  scale_fill_tableau(guide = "none") +
  facet_grid(plate_map_name ~ ccle_name, scales = 'free') +
  labs(x = "Replicate", y = "Spearman Correlation")
```

## Treatment dose trends

```{r, include = T, fig.width = 12, fig.height = 60}
normalized_data %>%
  dplyr::filter(pert_type == "treatment", !is.na(dose)) %>%
  ggplot(aes(x = dose, y = viability, color = arp_barcode)) +
  geom_line() + geom_point() +
  scale_x_log10() +
  facet_wrap(broad_id ~ ccle_name, scales = "free_x", ncol = 6) +
  labs(color = "Plate") +
  scale_color_tableau(palette = "Tableau 20") +
  ylim(c(-5, 150))
```
**Note:** these plots are limited to the range $[-5, 150]$ and exclude some outlying points (particularly negative values) on worse performing plates.


# Expected sensitivity compounds 

## Comparison with past screens

```{r comparison processing}
# validation compound info
val_compounds <- data.table::fread('val_compounds.csv', data.table = F) %>%
  .$pert_id
val_mapping <- data.table::fread('val_compounds.csv', data.table = F)

# read in old MTS files and clean up to have same columns (lazy fix)
old_mts_files <- list.files("old_processed_ctg_data", pattern = "*.Rds")
old_mts_data <- old_mts_files %>%
  purrr::set_names(str_sub(., 1, -5)) %>%
  purrr::map_dfr(~read_rds(file.path("old_processed_ctg_data", .x)) %>%
                   dplyr::select(ccle_name,
                                 any_of(c("brd_id", "broad_sample", "broad_id",
                                          "compound_plate", "plate_map_name",
                                          "dose", "pert_dose", "mmoles_per_liter",
                                          "pert_type", "trt_type")),
                                 source, viability, log_fc),
                 .id = "file") %>%
  dplyr::mutate(broad_sample = ifelse(is.na(broad_sample), brd_id, broad_sample),
                broad_sample = ifelse(is.na(broad_sample), broad_id, broad_sample),
                dose = ifelse(is.na(dose), pert_dose, dose),
                dose = ifelse(is.na(dose), mmoles_per_liter, dose),
                plate_map_name = ifelse(is.na(plate_map_name), compound_plate, plate_map_name),
                pert_type = ifelse(is.na(pert_type), trt_type, pert_type)) %>%
  dplyr::select(-brd_id, -broad_id, -pert_dose, -compound_plate, -trt_type, -mmoles_per_liter) %>%
  dplyr::mutate(pert_id = brd_id_to_structure_id(broad_sample),
                plate_map_name = paste0(plate_map_name, " (", source, ")")) %>%
  dplyr::rename(broad_id = broad_sample) %>%
  dplyr::mutate(pert_id = ifelse(str_detect(pert_id, fixed('BRD-A12230535')),
                                 "BRD-K73472992", pert_id))

# filter old data to validation only
old_mts_val <- old_mts_data %>% 
  dplyr::inner_join(val_mapping)

val_compound_ctg <- normalized_data %>%
  dplyr::mutate(pert_id = brd_id_to_structure_id(broad_id)) %>%
  dplyr::filter(pert_id %in% val_compounds)  %>%
  dplyr::mutate(plate_map_name = paste0(plate_map_name, " (", project, ")")) %>%
  dplyr::inner_join(val_mapping)
```

```{r}
# function to compare validation compounds of two screens
make_val_compound_plot <- function(new_df, old_df, comp_name, dataset) {
  
  # load in new and old screens
  new_df <- new_df %>%
    dplyr::filter(pert_name == comp_name) %>%
    dplyr::mutate(dose = as.numeric(dose))
  old_df <- old_df %>%
    dplyr::filter(source == dataset,
                  pert_name == comp_name) %>%
    dplyr::mutate(dose = as.numeric(dose)) %>%
    dplyr::filter(ccle_name %in% unique(new_df$ccle_name))
  
  # compute median viabilities
  new_median_viability <- new_df %>%
    dplyr::group_by(ccle_name, dose, plate_map_name, source) %>%
    dplyr::summarise(viability = median(viability, na.rm = T)) %>%
    dplyr::ungroup()
  old_median_viability <- old_df %>%
    dplyr::group_by(ccle_name, dose, plate_map_name, source) %>%
    dplyr::summarise(viability = median(viability, na.rm = T)) %>%
    dplyr::ungroup()
  
  # combine
  combined_viability <- dplyr::bind_rows(new_df, old_df)
  combined_median_viability <- dplyr::bind_rows(new_median_viability, old_median_viability)
  
  # fit curves
  curves <- combined_viability %>%
    dplyr::mutate(pert_effective_dose = as.numeric(dose), grouping = 1) %>%
    plyr::ddply(.(pert_name, plate_map_name, ccle_name), function(x) get_sigmoid(x, fitDoseCurve(x))) %>%
    dplyr::mutate(dose = exp(x), viability = y)
  
  # plot
  combined_viability %>%
    ggplot(aes(x = dose, y = viability, color = plate_map_name)) + 
    geom_point(alpha = .3) + 
    geom_point(data = new_df) +
    geom_line(data = dplyr::filter(curves, str_detect(plate_map_name, project))) +
    geom_line(data = curves, alpha = .3) +
    scale_x_log10() + 
    labs(x = "Dose", y = "Viability", color = "Plate") +
    facet_grid(. ~ ccle_name) + 
    scale_color_tableau() +
    ggtitle(paste(dataset, comp_name))
}

compounds <- distinct(old_mts_val, pert_name, source) %>%
  dplyr::filter(pert_name %in% unique(val_compound_ctg$pert_name)) %>%
  dplyr::arrange(desc(source))

comp_plots <- future_apply(compounds, 1,
                           FUN = function(x) make_val_compound_plot(val_compound_ctg, old_mts_val,
                                                                    x["pert_name"], x["source"]))
```

```{r comparison plots, fig.width = 10, fig.height = 4, include=T}
for(p in comp_plots) {
  print(p)
}
```

# Conclusion

We determined that the plates performed well enough to continue with the screen based on replicate correlation, SSMD, and dose-response behavior. There were two replicates (one of each plate) that performed poorly but were determined to be because of plating issues or CTG technical failure (because they were not from the same plate.)